### 2019-03-21 reports/thesis/chapters/5_inference.tex
Corrected setBandwidth in EmpiricalNull.m, added to inference chapter

The variance model, $\widehat{\variance}\left[\text{scan}_{i,j}\right]$, can be calibrated, or trained, by holding out a number replicated x-ray images of a 3D printed sample. These replicated x-ray images were manually segmented. This provided variance-mean data and was used to train the variance model, such as a Gamma distributed GLM. The method on doing so was described in the previous chapter and it was found a linear relationship between the variance and the mean was a good model. The variance was then predicted using the grey value in the \texttt{aRTist} simulation as the predictor variable.

@@

For each pixel, a $z$ statistic was calculated. A $z$ statistic too large in magnitude will classify that pixel as a positive result. In the hypothesis testing framework \citep{pearson1900on, neyman1933on, fisher1970statistical}, the $p$-value can be used to represent $z$ statistic in a different way
\begin{equation}
  p_{x,y} = 2(1-\Phi(|z_{x,y}|))
\end{equation}
which can takes values $0\leqslant p_{x,y} \leqslant 1$. A $p$-value too small is considered a positive result. The $p$-values are shown in Figure \ref{fig:inference_inferenceIntro_inference_pvalue}.

@@

\begin{figure}
  \centering
  \centerline{
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{../figures/inference/inferenceIntro_histogram.eps}
    \caption{Histogram of $z$}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{../figures/inference/inferenceIntro_pValue.eps}
    \caption{$p$-values}
  \end{subfigure}
}
  \caption{a) The histogram of the $z$ statistics is compared with the standard Normal distribution. The standard Normal distribution is known as the null distribution in this scenario. b) The $p$-values are ordered and plotted. The critical region corresponds to controlling the false discovery rate at 5\% using the \cite{benjamini1995controlling} procedure.}
  \label{fig:inference_inferenceIntro_histogram}
\end{figure}

This chapter recaps hypothesis testing, for a single test and then for multiple tests, treating each pixel as a test. Assumptions used on the hypothesis testing can be relaxed using the empirical null \citep{efron2004large}, which is reviewed here. The empirical null is then extended to an image filter, called the empirical null filter. This filter adjust each $z$ statistic according to its neighbours, ironing out false positive results. Simulations and results are shown towards the end of the chapter.

@@

In hypothesis testing \cite{pearson1900on, neyman1933on, fisher1970statistical}, the $p$-value can be compared with the user defined significance level $\alpha$, also known as the size of the test in this specific example. A positive result is declared when $p<\alpha$, else it is a negative result. Correct and incorrect testing of the statistics can occur. For example testing a null statistic as positive is known as a false positive. It can be shown that $\alpha$ controls the false positive rate. $\alpha=5.0\%$ is a typical choice and was used throughout this thesis. Other fields use the $\sigma$ notation, for example $2\sigma$ correspond to about the $\alpha = 4.6\%$ significance level; this notation will not be used here.

Multiple hypothesis testing occurs when there are more than one hypothesis to test. For example $n$ pixels can be tested where the $z$ statistics are $Z_1, Z_2, \cdots, Z_n$ and $Z_i\sim\normal(\mu_i,1)$ for $i=1,2,\cdots,n$. The null hypothesis is $H_{0,i}:\mu_i=0$ and is tested against the alternative $H_{1,i}:\mu_i\neq 0$ for $i=1,2,\cdots,n$. Let the corresponding $p$-values be $p_1, p_2, \cdots, p_n$.

@@

The BH procedure adjust the size of the test between the Bonferroni correction and the uncorrected test. It adapts to the data and choose different sizes for different data. The procedure is as follows, the $p$-values and ordered such that $p_{(1)}\leqslant p_{(2)}\leqslant \cdots \leqslant p_{(n)}$. Suppose a size $\alpha$ is provided beforehand. The size is then adjusted to
\begin{equation}
  \alpha_{\text{BH}} = \frac{\alpha k}{n}
\end{equation}
where
\begin{equation}
  k\text{ is the largest }i\text{ for which }p_{(i)}\leqslant\frac{i}{n}\alpha
  \ .
\end{equation}
This test the statistics with $p$-values $p_{(1)},p_{(2)},\cdots,p_{(k)}$ as positive. For the case where $p_{(1)}>\alpha/n$ then $k=1$ so that Bonferroni correction is used when there are no positive results, this is only for illustration purposes. The BH procedure comes from the fact that if all the statistics are null and independent, then the $p$-values are uniformly distributed \citep{simes1986improved}. However it can be shown that the BH procedure works for many scenarios of dependencies \citep{benjamini2001control}. It can be shown that the BH procedure controls the FDR such that
\begin{equation}
  \text{FDR}\leqslant\alpha
\end{equation}
\citep{benjamini1995controlling}.

@@

Consider a small example where a $200\times200$ section of the $z$ image was investigated, as shown Figure \ref{fig:inference_inferenceSubsample1_histogram}. The distribution of the $z$ statistics looked Normal but not centred at zero, as a result using the BH procedure obtained a critical region of $|Z|>\input{../figures/inference/inferenceSubsample1_criticalBoundary.txt}$. Almost half of the pixels were were tested positive. It is questionable whether such number of positive results is sensible, in particular in an area where defects were not expected.

@@

In \citep{efron2004large}, a few assumptions were made in order to obtain the estimators $\widehat{\mu}_0$ and $\widehat{\sigma}_0$. Let the test statistic $Z$ have the p.d.f.
\begin{equation}
  p_Z(z) =
  \pi_0 p_{Z|H_0}(z) + \pi_1 p_{Z|H_1}(z)
\end{equation}
where $0\leqslant\pi_0\leqslant 1$ and  $\pi_1 = 1-\pi_0$. In addition, the null distribution is
\begin{equation}
  p_{Z|H_0}(z) = 
  \dfrac{1}{\sqrt{2\pi}\widehat{\sigma}_0}
  \exp\left[
    -\dfrac{1}{2}
    \left(
      \dfrac{z-{\mu}_0}{{\sigma}_0}
    \right)^2
  \right]
\end{equation}
as it was assumed to be Normal. The non-null distribution $p_{Z|H_1}(z)$ does not need to be specified. Assume that the majority of the data are null and non-null test statistics are rare, say $\pi_0>0.9$ \citep{efron2004large}. Then around the mode, the probability density function would be dominated by the null distribution. This implies that
\begin{equation}
  p_{Z}(z) \approx \pi_0 p_{Z|H_0}(z)
\end{equation}
for values of $z$ around the mode. Finding the maximum for $p_{Z}(z)$ and $p_{Z|H_0}(z)$ should yield the same solution. This justifies the use of the mode for the empirical null mean \citep{efron2004large}
\begin{equation}
  \widehat{\mu}_0 = \argmax\widehat{p}_Z(z)
\end{equation}
where $\widehat{p}_Z(z)$ is the density estimation of $p_Z(z)$, for example it could be a smoothing spline fitted onto the histogram \citep{efron2004large}.

The null standard deviation was estimated from the log density \citep{efron2004large}. For values of $z$ at and around the mode
\begin{equation}
  \ln p_{Z}(z) = 
  \ln\left[
    \dfrac{\pi_0}{\sqrt{2\pi}{\sigma}_0}
  \right]
  -\dfrac{1}{2}
  \left(
    \dfrac{z-{\mu}_0}{{\sigma}_0}
  \right)^2
  \ .
\end{equation}
Taking derivatives
\begin{equation}
  \dfrac{\partial}{\partial z} \ln p_{Z}(z) =
  -\left(
    \dfrac{
      z-{\mu}_0
    }
    {
      {\sigma}_0^2
    }
  \right)
\end{equation}
\begin{equation}
  \dfrac{\partial^2}{\partial^2 z} \ln p_{Z}(z) =
  -\dfrac{
    1
  }
  {
    {\sigma}_0^2
  }
\end{equation}
which motivates the estimator \citep{efron2004large}
\begin{equation}
  \widehat{\sigma}_0 = \left[
    \left.
      -\dfrac{\partial^2}{\partial z^2}\ln\widehat{p}(z)
    \right|_{z=\widehat{\mu}_0}
  \right]^{-1/2}
  \ .
\end{equation}
The evaluation of $z=\widehat{\mu}_0$ is used because at the mode, it was assumed that the null test statistics would dominate and non-null test statistics would not contribute much to the density estimate.

@@

The empirical null is used widely, including in neuroimaging \citep{schwartzman2008false, schwartzman2009empirical} and has been extended to include non-Normal null distributions \citep{schwartzman2008false, schwartzman2008empirical}. There exist various methods for the empicial null. \citep{schwartzman2008empirical} uses Poisson regression on the histogram counts in regions where the null statistics dominate. The empirical characteristic function was used in \citep{jin2007estimating}.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.49\textwidth]{../figures/inference/inferenceSubsample1_densityEstimate.eps}
  \caption{The density of the $z$ statistics was estimated using a kernel density estimate. The dot dash lines shows the empirical null distribution.}
  \label{fig:inference_inferenceSubsample1_densityEstimate}
\end{figure}

@@

The mode was found by solving for $\widehat{\mu}_0 = \argmax\widehat{p}_Z(z)$. This cannot be solved in closed form easily and should be done numerically. The  Newton-Raphson method can be used to solve $\hat{p}_Z'(z) = 0$ for $z$. Instead the log space was used, solving
\begin{equation}
  \dfrac{
    \partial
  }
  {
    \partial z
  }
  \ln\widehat{p}_Z(z)
  = 0
\end{equation}
for $z$. This at least find stationary points.

@@

It was chosen that convergence was met when either 10 update steps were taken or when
\begin{equation}
  \log\left|
    \left.
    \dfrac{
      \partial
    }
    {
      \partial z
    }
  \ln\widehat{p}_Z(z)
  \right|_{z=\widehat{\mu}_0}
  \right|
  <-5
\end{equation}
at the current step. This was chosen arbitrary to speed up the algorithm without losing too much accuracy. At the end of the algorithm, for a successful convergence it was required, in addition, that
\begin{equation}
  \left.
    \dfrac{
      \partial^2
    }
    {
      \partial z^2
    }
    \ln\widehat{p}_Z(z)
  \right|_{z=\widehat{\mu}_0}
  < 0 \ .
\end{equation}
Following a successful convergence, the estimator $\widehat{\sigma}_0$ was calculated straight away.

@@

\subsection{Bandwidth Selection}

@@

In conclusion, a small bias added to the rule of thumb improved the performance of the estimator $\widehat{\sigma}_0$. \cite{silverman1986density} pointed out that a smaller bandwidth should be used when in particular the distribution is bimodal, this is why \cite{silverman1986density} suggest the use of $b=0.9$ rather than $b=1.06$. Following from this, it was chosen that the intercept was added to Silverman's rule of thumb for the use in the empirical null
\begin{equation}
  h = 0.9n^{-1/5}\times\text{min}\left(s_z,\text{IQR}_z/1.34\right) + 0.16
  \ .
  \tag{\ref{eq:inference_ourruleofthumb}}
\end{equation}

\section{Empirical Null Filter}

The empirical null is a good tool for defect detection under model misspecification. The empirical null could be used to solve the problem encountered at the start of the chapter. Recall that false positives were a problem due to model misspecification because the test statistics were not standard Normal distributed, as seen in Figure \ref{fig:inference_initial_artist_z_histo}. However the problem is that looking at the test statistics spatially in Figure \ref{fig:inference_initial_artist_logp_z}, the empirical null varied spatially. This is because it looked like the value of the test statistics depended on which surface it is on for example.

One possible solution is to spilt the $z$ image into a grid and conduct inference in each section separately. There are problems with this though. The first problem is that how the grid is overlaid can be arbitrary, for example the grid can be translated to produce different sections. Secondly it is not clear how to combine the results from each section \citep{efron2008simultaneous}.

The empirical null filter extends the empirical null to images. In this setting, the parameters of the null distribution varies spatially, slowly and smoothly in 2D space. Let $Z_{x,y}$ be the test statistic of the pixel at position $(x,y)$ for $x=1,2,\cdots,W$ and $y=1,2,\cdots,H$. Let each pixel have a test statistic distributed under the null such that
\begin{equation}
  Z_{x,y}|H_0\sim\normal\left(
    \mu_{x,y},\sigma_{0,x,y}^2
  \right)
  \ .
\end{equation}
Let the hypothesis
\begin{equation}
  H_0:\mu_{x,y}=\mu_{0,x,y}
\end{equation}
\begin{equation}
  H_1:\mu_{x,y}\neq\mu_{0,x,y}
\end{equation}
be tested for each pixel individually where $\mu_{0,x,y}$ and $\sigma_{0,x,y}$ are the null mean and null standard deviation at position $(x,y)$ respectively. Suppose $Z_{x,y}=z_{x,y}$ were observed for all $x$ and $y$. The empirical null filter aims to estimate the parameters $\mu_{0,x,y}$ and $\sigma_{0,x,y}$ for all $x$ and $y$. This is so that inference can be done under the empirical null distribution.

The empirical null filter uses the empirical null on neighbouring pixels to estimate the null distribution parameters. For example to estimate $\mu_{0,x,y}$ and $\sigma_{0,x,y}$, a circular kernel of radius $r$ is centred at $(x,y)$. All the pixels captured in the circular kernel are then used for the empirical null to obtain $\widehat{\mu}_{0,x,y}$ and $\widehat{\sigma}_{0,x,y}$, in other words
\begin{equation}
  \widehat{\mu}_{0,x,y}
  =
  \widehat{\mu}_{0,x,y}
    \left(
      \text{all }z_{i,j}\text{ such that }(x-i)^2+(y-j)^2\leqslant r^2
    \right)
\end{equation}
and
\begin{equation}
  \widehat{\sigma}_{0,x,y}
  =
  \widehat{\sigma}_{0,x,y}
    \left(
      \text{all }z_{i,j}\text{ such that }(x-i)^2+(y-j)^2\leqslant r^2
    \right)
  \ .
\end{equation}
Using these estimates, the test statistics are then corrected using
\begin{equation}
  \zeta_{x,y} = 
  \dfrac{
    z_{x,y}-\widehat{\mu}_{0,x,y}
  }
  {
    \widehat{\sigma}_{0,x,y}
  }
\end{equation}
and only the corrected test statistics are used in hypothesis testing.

The critical parameter of choice is the kernel radius $r$. A $r$ too small will treat defects as the null and will not detect them. A $r$ too big can also cause problems as well. The empirical null filter does assume the null parameters vary smoothly and slowly so that the null test statistics captured by the kernel is somewhat Normal. If $r$ is too big, the kernel will capture far away test statistics with a very different null distribution. If all of these test statistics are pulled together into a histogram, the null distribution may not look Normal at all which can cause problems when fitting a Normal null distribution. Furthermore, if $r\rightarrow \infty$, then this is just using the empirical null on all pixels which is, as discussed before, not sensible.

$r$ does control the proportion of pixels which are null captured by the kernel, denoted as $\pi_0$. If the size of the defects are known, then $r$ should be selected such that
\begin{equation}
\dfrac{
  \text{area of defect}
}
{
  \pi r^2
}
\approx
5\% \text{ to } 10 \%
\end{equation}
so that $\pi_0$ is of a sensible value. In practice, the defects are unknown so various kernel radiuses should be tried out. But in general, the kernel radius should be much larger than the size of the defect.

The empirical null filter was implemented in an open source software called \texttt{ImageJ} \citep{abramoff2004image, schneider2012nih, perez2013image} by modifying the existing class \texttt{RankFilters}, available on \texttt{GitHub} \citep{imagej2018rank}. This class implemented filters, such as the mean filter and median filter, using a circular kernel and multiple threads, making this a suitable framework for the empirical null filter. Figure \ref{fig:inference_fijiGui} shows the graphical user interface of the empirical null filter in \texttt{Fiji} \citep{schindelin2012fiji}, a distribution of \texttt{ImageJ}.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{../figures/inference/fiji/gui.png}
    \caption{Graphical user interface of the empirical null filter in \texttt{Fiji}. The user can adjust the kernel radius as well as other advanced options relating to the Newton-Raphson method and the kernel density estimate. By products such as the empirical null parameters can be shown after the filtering as well.}
    \label{fig:inference_fijiGui}
\end{figure}

As a test, the empirical null filter with $r=5$ was used on a \texttt{ImageJ} sample image \texttt{bridge.gif}, as shown in Figure \ref{fig:inference_fijiBridgeFilter}. It is meaningless to use the empirical null filter on an image not for the purpose of the filter, however it does verify the filter works on arbitrary images. The empirical null images $\widehat{\mu}_{0,x,y}$ and $\widehat{\sigma}_{0,x,y}$ for all ${x,y}$ could be of interest. In particular the empirical null mean image could be interpreted as the result of a mode filter \citep{griffin2000mean}. Figure \ref{fig:inference_fijiMean} compares the empirical null mean with other averaging filters. The empirical null mean creates an impasto look and preserve edges which is similar to the result in \cite{griffin2000mean}. Spread filters such as the standard deviation filter can be used to detect edges. Figure \ref{fig:inference_fijiStd} compares the standard deviation filter with the empirical null standard deviation. The resulting images are similar but it was notable that the edges in the empirical null standard deviation are sharper, suggesting that the empirical null standard deviation is a more robust measure of spread compared to the standard deviation.

>>>>

The variance model, $\widehat{\variance}\left[\text{scan}_{x,y}\right]$, can be calibrated, or trained, by holding out a number replicated x-ray images of a 3D printed sample. These replicated x-ray images were manually segmented. This provided variance-mean data and was used to train the variance model, such as a Gamma distributed GLM. The method on doing so was described in the previous chapter and it was found a linear relationship between the variance and the mean was a good model. The variance was then predicted using the grey value in the \texttt{aRTist} simulation as the predictor variable.

@@

For each pixel, a $z$ statistic was calculated. A $z$ statistic too large in magnitude will classify that pixel as a positive result. In two-tailed hypothesis testing framework \citep{pearson1900on, neyman1933on, fisher1970statistical}, the $p$-value can be used to represent $z$ statistic in a different way
\begin{equation}
  p_{x,y} = 2(1-\Phi(|z_{x,y}|))
\end{equation}
which can takes values $0\leqslant p_{x,y} \leqslant 1$. A $p$-value too small is considered a positive result. The $p$-values are shown in Figure \ref{fig:inference_inferenceIntro_inference_pvalue}.

@@

\begin{figure}
  \centering
  \centerline{
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{../figures/inference/inferenceIntro_histogram.eps}
    \caption{Histogram of $z$}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{../figures/inference/inferenceIntro_pValue.eps}
    \caption{$p$-values}
    \label{fig:inference_inferenceIntro_pvalue}
  \end{subfigure}
}
  \caption{a) The histogram of the $z$ statistics is compared with the standard Normal distribution. The standard Normal distribution is known as the null distribution in this scenario. b) The $p$-values are ordered and plotted. The critical region corresponds to controlling the false discovery rate at 5\% using the \cite{benjamini1995controlling} procedure.}
  \label{fig:inference_inferenceIntro_histogram}
\end{figure}

This chapter recaps hypothesis testing, for a single test and then for multiple tests, treating each pixel as a test. Assumptions used on the hypothesis testing can be relaxed by using the empirical null \citep{efron2004large}, which is reviewed here. The empirical null is then extended to an image filter, called the empirical null filter. This filter adjust each $z$ statistic according to its neighbours, ironing out false positive results. Simulations and results are shown towards the end of the chapter.

@@

In two-tailed hypothesis testing \cite{pearson1900on, neyman1933on, fisher1970statistical}, the $p$-value can be compared with the user defined significance level $\alpha$, also known as the size of the test in this specific example. A positive result is declared when $p<\alpha$, else it is a negative result. Correct and incorrect testing of the statistics can occur. For example testing a null statistic as positive is known as a false positive. It can be shown that $\alpha$ controls the false positive rate. $\alpha=5.0\%$ is a typical choice and was used throughout this thesis. Other fields use the $\sigma$ notation, for example $2\sigma$ correspond to about the $\alpha = 4.6\%$ significance level; this notation will not be used here.

Multiple hypothesis testing occurs when there are more than one hypothesis to test. For example $n$ pixels can be tested where the $z$ statistics are $Z_1, Z_2, \cdots, Z_n$ and $Z_i\sim\normal(\mu_i,1)$ for $i=1,2,\cdots,n$. The null hypotheses are $H_{0,i}:\mu_i=0$ and are tested against the alternative $H_{1,i}:\mu_i\neq 0$ for $i=1,2,\cdots,n$. Let the corresponding $p$-values be $p_1, p_2, \cdots, p_n$.

@@

The BH procedure adjust the size of the test between the Bonferroni correction and the uncorrected test. It adapts to the data and choose different sizes for different data. The procedure is as follows, the $p$-values are ordered such that $p_{(1)}\leqslant p_{(2)}\leqslant \cdots \leqslant p_{(n)}$. Suppose a size $\alpha$ is provided beforehand. The size is then adjusted to
\begin{equation}
  \alpha_{\text{BH}} = \frac{\alpha k}{n}
\end{equation}
where
\begin{equation}
  k\text{ is the largest }i\text{ for which }p_{(i)}\leqslant\frac{i}{n}\alpha
  \ .
\end{equation}
This test the statistics with $p$-values $p_{(1)},p_{(2)},\cdots,p_{(k)}$ as positive. For the case where $p_{(1)}>\alpha/n$ then $k=1$ so that Bonferroni correction is used when there are no positive results, this is only for illustration purposes. For example in Figure \ref{fig:inference_inferenceIntro_pvalue} the $p$-values were plotted against their order so that the critical boundary was shown by a linear curve with gradient $\alpha/n$.

The BH procedure comes from the fact that if all the statistics are null and independent, then the $p$-values are uniformly distributed \citep{simes1986improved}. However it can be shown that the BH procedure works for many scenarios of dependencies \citep{benjamini2001control}. It can be shown that the BH procedure controls the FDR such that
\begin{equation}
  \text{FDR}\leqslant\alpha
\end{equation}
\citep{benjamini1995controlling}.

@@

Consider a small example where a $200\times200$ section of the $z$ image was investigated, as shown Figure \ref{fig:inference_inferenceSubsample1_histogram}. The distribution of the $z$ statistics looked Normal but not centred at zero, as a result using the BH procedure obtained a critical region of $|Z|>\input{../figures/inference/inferenceSubsample1_criticalBoundary.txt}$. Almost half of the pixels were tested positive. It is questionable whether such a number of positive results is sensible, in particular in an area where defects were not expected.

@@

In \cite{efron2004large}, a few assumptions were made in order to obtain the estimators $\widehat{\mu}_0$ and $\widehat{\sigma}_0$. Let the test statistic $Z$ have the p.d.f.
\begin{equation}
  p_Z(z) =
  \pi_0 p_{Z|H_0}(z) + \pi_1 p_{Z|H_1}(z)
\end{equation}
where $0\leqslant\pi_0\leqslant 1$ and  $\pi_1 = 1-\pi_0$. In addition, the null distribution is
\begin{equation}
  p_{Z|H_0}(z) = 
  \dfrac{1}{\sqrt{2\pi}\sigma_0}
  \exp\left[
    -\dfrac{1}{2}
    \left(
      \dfrac{z-{\mu}_0}{\sigma_0}
    \right)^2
  \right]
\end{equation}
as it was assumed to be Normal. The non-null distribution $p_{Z|H_1}(z)$ does not need to be specified. Assume that the majority of the data are null and non-null test statistics are rare, say $\pi_0>0.9$ \citep{efron2004large}. Then around the mode, the probability density function would be dominated by the null distribution. This implies that
\begin{equation}
  p_{Z}(z) \approx \pi_0 p_{Z|H_0}(z)
\end{equation}
for values of $z$ around the mode. Finding the mode for $p_{Z}(z)$ and $p_{Z|H_0}(z)$ should yield the same solution. This justify the use of the mode for the empirical null mean \citep{efron2004large}
\begin{equation}
  \widehat{\mu}_0 = \argmax\widehat{p}_Z(z)
\end{equation}
where $\widehat{p}_Z(z)$ is the density estimation of $p_Z(z)$, for example it could be a smoothing spline fitted onto the histogram \citep{efron2004large}.

The null standard deviation was estimated from the log density \citep{efron2004large}. For values of $z$ at and around the mode
\begin{equation}
  \ln p_{Z}(z) = 
  \ln\left[
    \dfrac{\pi_0}{\sqrt{2\pi}{\sigma}_0}
  \right]
  -\dfrac{1}{2}
  \left(
    \dfrac{z-{\mu}_0}{{\sigma}_0}
  \right)^2
  \ .
\end{equation}
Taking derivatives
\begin{equation}
  \dfrac{\partial}{\partial z} \ln p_{Z}(z) =
  -\left(
    \dfrac{
      z-{\mu}_0
    }
    {
      {\sigma}_0^2
    }
  \right)
\end{equation}
\begin{equation}
  \dfrac{\partial^2}{\partial z^2} \ln p_{Z}(z) =
  -\dfrac{
    1
  }
  {
    {\sigma}_0^2
  }
\end{equation}
which motivates the estimator \citep{efron2004large}
\begin{equation}
  \widehat{\sigma}_0 = \left[
    \left.
      -\dfrac{\partial^2}{\partial z^2}\ln\widehat{p}(z)
    \right|_{z=\widehat{\mu}_0}
  \right]^{-1/2}
  \ .
\end{equation}
The evaluation of $z=\widehat{\mu}_0$ is used because at the mode, it was assumed that the null test statistics would dominate and non-null test statistics would not contribute much to the density estimate.

@@

The empirical null is used widely, including in neuroimaging \citep{schwartzman2008false, schwartzman2009empirical} and has been extended to include non-Normal null distributions \citep{schwartzman2008false, schwartzman2008empirical}. There exist various methods for the empicial null. \cite{schwartzman2008empirical} uses Poisson regression on the histogram counts in regions where the null statistics dominate. The empirical characteristic function was used in \cite{jin2007estimating}.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.49\textwidth]{../figures/inference/inferenceSubsample1_densityEstimate.eps}
  \caption{The density of the $z$ statistics was estimated using a kernel density estimate. The dot dash lines shows the empirical null distribution multiplied by some constant, this is to insulate the two densities have the same curvature at the mode.}
  \label{fig:inference_inferenceSubsample1_densityEstimate}
\end{figure}

@@

The mode was found by solving $\widehat{\mu}_0 = \argmax\widehat{p}_Z(z)$. This cannot be solved in closed form easily and was done numerically. The  Newton-Raphson method can be used to solve $\hat{p}_Z'(z) = 0$ for $z$. Instead the log space was used, solving
\begin{equation}
  \dfrac{
    \partial
  }
  {
    \partial z
  }
  \ln\widehat{p}_Z(z)
  = 0
\end{equation}
for $z$. This at least find stationary points.

@@

A convergence criteria was met when either 10 update steps were taken or when
\begin{equation}
  \log\left[\left|
    \left.
    \dfrac{
      \partial
    }
    {
      \partial z
    }
  \ln\widehat{p}_Z(z)
  \right|_{z=\widehat{\mu}_0}
  \right|\right]
  <-5
\end{equation}
at the current step. This was chosen arbitrary to speed up the algorithm without losing too much accuracy. At the end of the algorithm, for a successful convergence it was required, in addition, that
\begin{equation}
  \left.
    \dfrac{
      \partial^2
    }
    {
      \partial z^2
    }
    \ln\widehat{p}_Z(z)
  \right|_{z=\widehat{\mu}_0}
  < 0 \ .
\end{equation}
Following a successful convergence, the estimator $\widehat{\sigma}_0$ was calculated straight away.

@@

\subsection{Bandwidth Tuning}

@@

In conclusion, a small bias added to the rule of thumb improved the performance of the estimator $\widehat{\sigma}_0$. \cite{silverman1986density} pointed out that a smaller bandwidth should be used when in particular the distribution is bimodal, this is why \cite{silverman1986density} suggest the use of $b=0.9$ rather than $b=1.06$. Following from this, it was chosen that the intercept was added to Silverman's rule of thumb for the use in the empirical null
\begin{equation}
  h = \left(0.9n^{-1/5}+ 0.16\right) \times\text{min}\left(s_z,\text{IQR}_z/1.34\right) 
  \ .
  \tag{\ref{eq:inference_ourruleofthumb}}
\end{equation}

\afterpage{\clearpage}
\subsection{Sampling Distribution Experiments}

Various estimators of central tendency and dispersion were compared with the empirical null estimates of the null mean and null standard deviation. The empirical sampling distribution of these estimators were observed to look for properties such as bias, variance and robustness. The central tendency looked at were the median and mean. The dispersions looked at were the median absolute deviation (MAD) around the mode $\times 1.4826$, interquartile range (IQR) $\div 1.349$ and the standard deviation. The normalisation constants were used so that the MAD and IQR can be compared directly to the standard deviation.

A dataset of $n = \pi r^2$ standard Normal random variables were simulated to obtain estimates of central tendency and dispersion. This was repeated 100 times to obtain a sampling distribution. The results are shown in Figures \ref{fig:inference_NullIidMean} and \ref{fig:inference_NullIidStd}. The null mean and standard deviation estimations both suffer from high variance, much higher than typical robust estimators such as the median and IQR.

The estimation of the null standard deviation has some positive bias and it was found that this bias is quite sensitive to the bandwidth. Further fine tuning on the bandwidth would be required for the empirical null standard deviation to be unbiased. For example it appeared there is consistent positive bias for all $r$ investigated, perhaps suggesting the bandwidth parameter $a$ should be decreased ever so slightly. The source of bias could come from the fact that the bandwidth was tuned to minimise the mean squared error rather than the squared bias.

To test for robustness, a dataset of $n = \pi r^2$ mixture of Gaussian random variables were simulated with the following distribution: $\normal(0,1)$ with probability $90\%$, $\normal(3,1)$ with probability $10\%$. The $\normal(3,1)$ component acts as non-null statistics and robust estimators should be unaffected by it. The results are shown in Figures \ref{fig:inference_NullIidMixtureMean} and \ref{fig:inference_NullIidMixtureStd}. The empirical null estimators are robust to the non-null component but still suffers from high variance. The rest of the estimators were affected by the non-null statistics, making them unsuitable to estimate parameters of the null component.

\begin{figure}[p]
  \centering
    \centerline{
      \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{../figures/inference/NullIid/NullIidEmpirical_nullMean.eps}
        \caption{Empirical null mean}
      \end{subfigure}
    }
    \centerline{
      \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{../figures/inference/NullIid/NullIidMedianIqr_nullMean.eps}
        \caption{Median}
      \end{subfigure}
      \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{../figures/inference/NullIid/NullIidMeanVar_nullMean.eps}
        \caption{Mean}
      \end{subfigure}
    }
  \caption{Different types of central tendency estimators were used on $\pi r^2$ standard Normal random variables. Shown are the sampling distributions from 100 repeats. The dotted lines show the 95\% confidence interval for the sample mean.}
  \label{fig:inference_NullIidMean}
\end{figure}

\begin{figure}
  \centering
    \centerline{
      \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{../figures/inference/NullIid/NullIidEmpirical_nullStd.eps}
        \caption{Empirical null std.}
      \end{subfigure}
      \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{../figures/inference/NullIid/NullIidMadMode_nullStd.eps}
        \caption{MAD mode}
      \end{subfigure}
    }
    \centerline{
      \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{../figures/inference/NullIid/NullIidMedianIqr_nullStd.eps}
        \caption{IQR}
      \end{subfigure}
      \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{../figures/inference/NullIid/NullIidMeanVar_nullStd.eps}
        \caption{Standard deviation}
      \end{subfigure}
    }
  \caption{Different types of statistical dispersion estimators were used on $\pi r^2$ standard Normal random variables. Shown are the sampling distributions from 100 repeats. The dotted lines show the 95\% confidence interval for the sample variance.}
  \label{fig:inference_NullIidStd}
\end{figure}

\begin{figure}
  \centering
    \centerline{
      \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{../figures/inference/NullIid/NullIidMixtureEmpirical_nullMean.eps}
        \caption{Empirical null mean}
      \end{subfigure}
    }
    \centerline{
      \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{../figures/inference/NullIid/NullIidMixtureMedianIqr_nullMean.eps}
        \caption{Median}
      \end{subfigure}
      \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{../figures/inference/NullIid/NullIidMixtureMeanVar_nullMean.eps}
        \caption{Mean}
      \end{subfigure}
    }
  \caption{Different types of central tendency estimators were used on $\pi r^2$ mixture of Gaussian random variables. The random variable has distribution $\normal(0,1)$ with probability $0.9$ and $\normal(3,1)$ with probability $0.1$. Shown are the sampling distributions from 100 repeats. The dotted lines show the 95\% confidence interval for the sample mean.}
  \label{fig:inference_NullIidMixtureMean}
\end{figure}

\begin{figure}[p]
  \centering
    \centerline{
      \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{../figures/inference/NullIid/NullIidMixtureEmpirical_nullStd.eps}
        \caption{Empirical null std.}
      \end{subfigure}
      \begin{subfigure}[b]{0.49\textwidth}
        \includegraphics[width=\textwidth]{../figures/inference/NullIid/NullIidMixtureMadMode_nullStd.eps}
        \caption{MAD mode}
      \end{subfigure}
    }
    \centerline{
      \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{../figures/inference/NullIid/NullIidMixtureMedianIqr_nullStd.eps}
        \caption{IQR}
      \end{subfigure}
      \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{../figures/inference/NullIid/NullIidMixtureMeanVar_nullStd.eps}
        \caption{Standard deviation}
      \end{subfigure}
    }
  \caption{Different types of statistical dispersion estimators were used on $\pi r^2$ mixture of Gaussian random variables. The random variable has distribution $\normal(0,1)$ with probability $0.9$ and $\normal(3,1)$ with probability $0.1$. Shown are the sampling distributions from 100 repeats. The dotted lines show the 95\% confidence interval for the sample variance.}
  \label{fig:inference_NullIidMixtureStd}
\end{figure}

\afterpage{\clearpage}
\section{Empirical Null Filter}

Returning back to the example at the start of the chapter, the empirical null could be used on the histogram shown in Figure \ref{fig:inference_inferenceIntro_histogram}. The resulting empirical null distribution is shown in Figure \ref{fig:inference_inferenceSubsampleAll}. The empirical null distribution has a very large standard deviation and would test all pixels as negative, this is incorrect.
